---
format:
  revealjs: 
    theme: [default, hadley.scss]
---

# Data manipulation with dplyr

## Getting started

```{r}
library(tidyverse)
library(palmerpenguins)
```

## Big picture

-   dplyr provides a selection of verbs that all take a data frame as input, do something to it and return a modified data frame.
-   Each verb is simple in isolation; you solve complex problems by chaining them together with the pipe (`|> in`base R \>= 4.1 or `%>% from magrittr/tidyverse`).
-   Learn more in <https://r4ds.hadley.nz/data-transform.html>.

## Selecting columns with `select()`

-   We're not going to cover it because `penguins` already has a small set of columns.
-   But `select()` is super useful when you have a very wide dataset and want to zoom in on the variables you care about.

## Filtering rows with `filter()`

-   `filter(df, condition)` returns all rows where `condition` evaluates to `TRUE`
-   How do you make `TRUE`s and `FALSE`s? (collectively called a logical vector)
    -   Compare: `body_mass_g < 3000`, `species == "Chinstrap"`
    -   Find one of multiple options: `island %in% c("Dream", "Torgersen")`
    -   Combine conditions with logical operators: `&` (and), `|` (or)
    -   Negate a condition: `!` (not)
-   More options at <https://r4ds.hadley.nz/logicals.html>. 

## Your turn

-   Find all measurements made after 2007
-   Find all male Chinstrap penguins
-   Find all penguins that have a bill over 50 mm long or weight more than 5kg.

## Common problems

-   Console shows `+` and nothing is happening --- you're missing a `"` or a `)`. Easiest to press `Esc` and start again
-   We detected a named input --- R uses `==` for comparison, not `=`
-   What else?

## Missing values

What's going on?

```{r}
filter(penguins, sex == NA)
```

Easier to understand with a concrete example:

```{r}
hadley_height <- NA
jenny_height <- NA
hadley_height == jenny_height
```

## The solution: `is.na()`

```{r}
filter(penguins, is.na(sex))
```

## Sorting the rows

-   Use `arrange(df, variable)` to sort by `variable`, from smallest to largest.
-   Use `arrange(df, desc(variable))` to sort by `variable`, from largest to smallest.
-   Use `arrange(df, var1, var2)` to sort by `var1` then `var2`

## Your turn

-   Find the heaviest penguins
-   Find the penguins with the shortest bills

## Filtering and sorting

-   What happens if you want to filter AND sort?
-   You need the pipe `%>%` or `|>`.
-   Pronounced "then"
-   Easiest way to type is to use the `Cmd + shift + m` shortcut
-   Use little bunny foofoo or cupcakes to explain the point?

------------------------------------------------------------------------

```{r}
penguins |> 
  filter(sex == "male") |> 
  arrange(desc(body_mass_g))

# Can also use with ggplot2!
penguins |> 
  filter(sex == "male") |>
  ggplot(aes(bill_length_mm)) +
  geom_histogram(binwidth = 0.5)
```

## Adding new variables

-   `mutate(df, new_var = computation)` creates a new variable called `new_var` from the `computation`
-   Can create multiple variables in one step.
-   Arithmetic: `+`, `*`, `-`, `/`, `^`
-   Conditional: `if_else(logical_condition, true, false)`
-   Many other options in <https://r4ds.hadley.nz/numbers.html>

------------------------------------------------------------------------

```{r}
penguins |> 
  mutate(
    bill_ratio = bill_length_mm / bill_depth_mm,
    size = if_else(body_mass_g > 400, "big", "small")
  )
```

## Uhoh

-   We can't see the columns we just created
-   Options:
    -   Get a bigger screen 😆

    -   `|> View()`

    -   `.keep = "used"` or "unused"?

    -   no love for `.before` and `.after`?
        I like them

## Your turn

-   Create `bill_length_cm`, `bill_depth_cm`, and `body_mass_kg`
-   Create a new `bill_type` variable that is "skinny" if `bill_ratio` \< 0.5 and "stumpy" if not.
-   Explore the different ways to view the computation

## Creating objects

```{r, error = TRUE}
penguins |> 
  mutate(
    bill_ratio = bill_length_mm / bill_depth_mm,
    size = if_else(body_mass_g > 400, "big", "small")
  )

penguins |> 
  arrange(bill_ratio)
```

------------------------------------------------------------------------

dplyr functions return a new data frame.
They never modify the input or automatically save the output.
You need to save yourself with `<-`:

```{r}
penguins2 <- penguins |> 
  mutate(
    bill_ratio = bill_length_mm / bill_depth_mm,
    size = if_else(body_mass_g > 400, "big", "small")
  )

penguins2 |> 
  arrange(bill_ratio)
```

## Grouped summaries

```{r}
penguins |> count(species)
penguins |> count(island)
penguins |> count(species, island)
```

How do we do this for our own summaries?

------------------------------------------------------------------------

```{r}
penguins |> 
  group_by(island) |> 
  summarise(n = n())
```

## Summary options

-   `n()`, `n_distinct(variable)`
-   `mean(variable)`, `sd(variable)`
-   `min(variable)`, `median(variable)`, `quantile(variable, 0.75)`, `max(variable)`
-   <https://r4ds.hadley.nz/numbers.html#summaries>

## NA values strike again!

```{r}
penguins |> 
  group_by(island) |> 
  summarise(body_mass_g = mean(body_mass_g))

penguins |> 
  group_by(island) |> 
  summarise(body_mass_g = mean(body_mass_g, na.rm = TRUE))
```

## Your turn

-   How many species are there on each island? How many islands does each species occupy?
-   How does the average bill size of each species vary?
-   Compare the 25th and 75th quantile of `body_mass_g` for each species.

## Grouped mutates

```{r}
penguins |> 
  group_by(species) |> 
  filter(!is.na(body_mass_g)) |> 
  mutate(
    body_mass_z = (body_mass_g - mean(body_mass_g)) / sd(body_mass_g)
  ) |> 
  ggplot(aes(body_mass_z, colour = species)) +
  geom_freqpoly(binwidth = 0.25)
```

## Where next with dplyr?

-   Other functions like `select()`, `distinct()`, `rename()`, `relocate()`, `slice_*`
-   Functions that work with multiple tables: `left_join()`, `semi_join()`, ...
-   Other types of variables: strings, factors, dates, ...

## Where next with the tidyverse?

-   Data import: csv (**readr**), SAS/SPSS/Stata (**haven**), spreadsheets (**readxl**, **googlesheets4**), databases (**dbplyr**), HTML (**rvest**), ...
-   Data tidying: get your variables into columns and observations into rows with tidyr.
-   Communicating: use quarto to mingle prose and code and produce docs, presentations, websites, books, ...

# Thank you!
